---
title: "Dynamics of aquatic and soil-derived bacteria in a temperate lake"
author: "Tajinder Singh, Nathan I. Wisnoski"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
header-includes:
- \usepackage{array}
- \usepackage{graphics}
---

# Overview

This analysis evaluates the dynamics of aquatic and soil-derived bacterial OTUs in a temperate lake time series. The workflow identifies OTUs detected in soil samples, evaluates whether those OTUs are also present or active in aquatic transect and time-series samples, classifies temporal activity patterns, and summarizes their contribution to the active aquatic community.

# Setup

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

## Load packages and helper functions

```{r packages, results='hide'}
library(tidyverse)
library(vegan)
library(viridis)
library(lubridate)
library(readxl)
library(writexl)
library(colorspace)
library(ggpubr)

source("bin/mothur_tools.R")

se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(na.omit(x)))
}

cv <- function(x) {
  sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
}
```

## Define paths

```{r paths}
paths <- list(
  design = "data/UL.design.txt",
  shared = "data/ul_resgrad.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.opti_mcc.shared",
  taxonomy = "data/ul_resgrad.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.opti_mcc.0.03.cons.taxonomy",
  timeseries_design = "data/UL_timeseries_design.csv",
  env_data = "data/ul-seedbank.env.txt",
  figures = "figures"
)

dir.create(paths$figures, showWarnings = FALSE, recursive = TRUE)
```

## Figure theme

```{r figure_theme}
grey_cols <- c(
  "grey60",
  "grey20"
)

activity_colors <- c(
  transient   = "#D55E00",
  seasonal    = "#009E73",
  persistent  = "#0072B2",
  unpatterned = "#999999"
)

abundance_colors <- c(
  aggregate_abundance = "#4D4D4D",
  common_abundance    = "#0072B2",
  rare_abundance      = "#CC79A7"
)

activity_linetypes <- c(
  transient   = "dotdash",
  seasonal    = "dashed",
  persistent  = "solid",
  unpatterned = "dotted"
)

abundance_linetypes <- c(
  aggregate_abundance = "solid",
  common_abundance = "dashed",
  rare_abundance = "dotdash"
)

theme_set(
  theme_classic() +
    theme(
      axis.title = element_text(size = 14),
      axis.title.x = element_text(margin = margin(t = 15, b = 15)),
      axis.title.y = element_text(margin = margin(l = 15, r = 15)),
      axis.text = element_text(size = 12),
      axis.text.x = element_text(margin = margin(t = 5)),
      axis.text.y = element_text(margin = margin(r = 5)),
      axis.line.x = element_blank(),
      axis.line.y = element_blank(),
      axis.ticks.x = element_line(linewidth = 1),
      axis.ticks.y = element_line(linewidth = 1),
      axis.ticks.length = unit(.1, "in"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
      legend.title = element_blank(),
      legend.text = element_text(size = 14),
      strip.text = element_text(size = 12),
      strip.background = element_blank()
    )
)
```

# Helper functions

```{r helper_functions}
make_presence <- function(x) {
  decostand(x, method = "pa")
}

make_relative_abundance <- function(x) {
  decostand(x, method = "total")
}

make_long_otu_table <- function(otu_table, metadata, value_name = "relative_abundance") {
  otu_table |>
    as_tibble() |>
    mutate(sample_id = as.integer(metadata$sample.id), .before = 1) |>
    pivot_longer(
      cols = -sample_id,
      names_to = "OTU",
      values_to = value_name
    ) |>
    left_join(metadata, by = c("sample_id" = "sample.id"))
}

summarize_taxon_abundance <- function(long_data, taxa, env_data) {
  long_data |>
    filter(OTU %in% taxa) |>
    group_by(sample_id) |>
    summarize(relative_abundance = sum(relative_abundance), .groups = "drop") |>
    left_join(env_data |> select(sample.id, date), by = c("sample_id" = "sample.id"))
}

summarize_common_rare_abundance <- function(long_data, taxa, total_abundance) {
  common_abundance <-
    long_data |>
    filter(OTU %in% taxa, abundance_type == "common") |>
    group_by(sample_id) |>
    summarize(common_abundance = sum(relative_abundance), .groups = "drop")

  rare_abundance <-
    long_data |>
    filter(OTU %in% taxa, abundance_type == "rare") |>
    group_by(sample_id) |>
    summarize(rare_abundance = sum(relative_abundance), .groups = "drop")

  total_abundance |>
    rename(aggregate_abundance = relative_abundance) |>
    left_join(common_abundance, by = "sample_id") |>
    left_join(rare_abundance, by = "sample_id") |>
    mutate(across(contains("abundance"), ~ replace_na(.x, 0)))
}

plot_presence_facets <- function(long_presence_data, taxa, x_var = date, color = NULL) {
  # Retained only for diagnostic checks of occupancy-based classifications.
  # Main visual interpretation should use relative-abundance plots.
  p <-
    long_presence_data |>
    filter(OTU %in% taxa) |>
    ggplot(aes(x = {{ x_var }}, y = presence, group = OTU)) +
    geom_line(alpha = 0.3) +
    facet_wrap(~OTU) +
    scale_y_continuous(limits = c(-0.5, 1.5)) +
    theme_minimal()

  if (!is.null(color)) {
    p <- p + geom_line(alpha = 1, color = color)
  }

  p
}

plot_relative_abundance_facets <- function(long_data, taxa, x_var = date) {
  long_data |>
    filter(OTU %in% taxa) |>
    ggplot(aes(x = {{ x_var }}, y = relative_abundance, group = OTU)) +
    geom_line(alpha = 0.45, linewidth = 0.35, color = "grey45") +
    facet_wrap(~OTU, scales = "free_y") +
    labs(x = "Date", y = "Relative abundance") +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(size = 8),
      axis.text = element_text(size = 7),
      panel.grid.minor = element_blank()
    )
}

make_timeseries_long <- function(otu_table, metadata, molecule_type) {
  otu_table |>
    as_tibble() |>
    mutate(sample_id = as.integer(metadata$sample.id), .before = 1) |>
    pivot_longer(
      cols = -sample_id,
      names_to = "OTU",
      values_to = "relative_abundance"
    ) |>
    left_join(metadata, by = c("sample_id" = "sample.id")) |>
    mutate(molecule_type = molecule_type)
}

plot_example_dna_rna_trajectories <- function(dna_long, rna_long, example_taxa) {
  bind_rows(dna_long, rna_long) |>
    filter(OTU %in% example_taxa$OTU) |>
    left_join(example_taxa, by = "OTU") |>
    mutate(
      activity_pattern = factor(activity_pattern, levels = c("persistent", "seasonal", "transient")),
      molecule_type = factor(molecule_type, levels = c("DNA", "RNA"))
    ) |>
    ggplot(aes(x = date, y = relative_abundance, color = activity_pattern)) +
    geom_line(linewidth = 0.7, alpha = 0.85, show.legend = FALSE) +
    geom_vline(xintercept = as_date("2013-07-01")) +
    facet_grid(activity_pattern + OTU ~ molecule_type, scales = "free_y") +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
    scale_color_manual(values = activity_colors, drop = FALSE) +
    labs(
      x = "",
      y = "Relative abundance",
      color = "Activity pattern"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}

get_max_relative_abundance <- function(otu_table, prefix) {
  otu_table |>
    as.data.frame() |>
    make_relative_abundance() |>
    t() |>
    as.data.frame() |>
    filter(rowSums(pick(everything())) > 0) |>
    mutate(max_relative_abundance = apply(pick(everything()), 1, max)) |>
    mutate(max_abundance_type = if_else(max_relative_abundance > 0.001, "common", "rare")) |>
    rownames_to_column(var = "OTU") |>
    select(OTU, max_relative_abundance, max_abundance_type) |>
    rename(
      "{prefix}_max_relative_abundance" := max_relative_abundance,
      "{prefix}_max_abundance_type" := max_abundance_type
    )
}

classify_activity_patterns <- function(
  long_data,
  presence_table,
  taxa = colnames(presence_table),
  date_col = date,
  abundance_col = relative_abundance,
  persistent_cutoff = 0.90,
  transient_cutoff = 0.10,
  seasonal_min_years = 2,
  seasonal_min_peak_months = 2,
  seasonal_max_peak_months = 6,
  seasonal_concentration_cutoff = 0.35,
  seasonal_recurrence_cutoff = 0.40,
  seasonal_min_presence_frequency = 0.05,
  seasonal_min_max_run = 2,
  seasonal_min_peak_window_active_samples = 4,
  seasonal_peak_to_background_cutoff = 3,
  seasonal_relative_amplitude_cutoff = 0.50,
  seasonal_abundance_peak_to_background_cutoff = 3,
  seasonal_abundance_relative_amplitude_cutoff = 0.50,
  seasonal_peak_doy_concentration_cutoff = 0.60,
  seasonal_min_peak_years = 2,
  seasonal_max_single_year_peak_fraction = 0.75,
  transient_single_year_dominance_cutoff = 0.75,
  transient_peak_to_background_cutoff = 3,
  seasonal_harmonic_r2_cutoff = 0.10,
  seasonal_harmonic_amplitude_cutoff = 0.25,
  smooth_activity = TRUE,
  smooth_window = 3,
  abundance_floor = 0
) {
  # Classifies OTU activity patterns using occupancy, recurrence, run length,
  # and seasonal concentration.
  #
  # Definitions:
  # - persistent: active in most samples.
  # - transient: active in few samples.
  # - seasonal: intermediate occupancy, recurring across years, and concentrated
  #   in a limited part of the annual cycle.
  # - unpatterned: intermediate occupancy without consistent seasonal structure.
  #
  # Smoothing is optional and conservative. It is used only to stabilize the
  # seasonal signal, not to redefine whether a taxon was observed in a sample.

  stopifnot(all(taxa %in% colnames(presence_table)))

  presence_table <- presence_table[, taxa, drop = FALSE]

  metadata <-
    long_data |>
    distinct(sample_id, {{ date_col }}) |>
    arrange({{ date_col }}) |>
    mutate(
      year = lubridate::year({{ date_col }}),
      month = lubridate::month({{ date_col }}),
      day_of_year = lubridate::yday({{ date_col }}),
      theta = 2 * pi * (day_of_year - 1) / 365.25
    )

  if (nrow(metadata) != nrow(presence_table)) {
    stop("presence_table and long_data metadata do not have the same number of samples.")
  }

  presence_long <-
    presence_table |>
    as_tibble() |>
    mutate(sample_id = metadata$sample_id, .before = 1) |>
    pivot_longer(
      cols = -sample_id,
      names_to = "OTU",
      values_to = "presence"
    ) |>
    left_join(metadata, by = "sample_id") |>
    filter(OTU %in% taxa)

  abundance_long <-
    long_data |>
    filter(OTU %in% taxa) |>
    select(sample_id, OTU, abundance = {{ abundance_col }})

  taxon_time_series <-
    presence_long |>
    left_join(abundance_long, by = c("sample_id", "OTU")) |>
    mutate(
      abundance = replace_na(abundance, 0),
      active_observed = presence == 1 & abundance >= abundance_floor
    ) |>
    arrange(OTU, {{ date_col }}) |>
    group_by(OTU) |>
    mutate(
      active_smoothed = if (smooth_activity) {
        as.numeric(stats::filter(active_observed, rep(1 / smooth_window, smooth_window), sides = 2))
      } else {
        as.numeric(active_observed)
      },
      active_smoothed = coalesce(active_smoothed, as.numeric(active_observed)),
      abundance_smoothed = if (smooth_activity) {
        as.numeric(stats::filter(abundance, rep(1 / smooth_window, smooth_window), sides = 2))
      } else {
        abundance
      },
      abundance_smoothed = coalesce(abundance_smoothed, abundance)
    ) |>
    ungroup()

  run_summary <-
    taxon_time_series |>
    arrange(OTU, {{ date_col }}) |>
    group_by(OTU) |>
    mutate(run_id = cumsum(active_observed != lag(active_observed, default = first(active_observed)))) |>
    group_by(OTU, run_id, active_observed) |>
    summarize(run_length = n(), .groups = "drop") |>
    filter(active_observed) |>
    group_by(OTU) |>
    summarize(
      max_active_run = max(run_length, na.rm = TRUE),
      mean_active_run = mean(run_length, na.rm = TRUE),
      .groups = "drop"
    )

  annual_summary <-
    taxon_time_series |>
    group_by(OTU, year) |>
    summarize(
      annual_occupancy = mean(active_observed, na.rm = TRUE),
      detected_this_year = any(active_observed),
      .groups = "drop"
    ) |>
    group_by(OTU) |>
    summarize(
      years_detected = sum(detected_this_year),
      years_sampled = n(),
      annual_recurrence = years_detected / years_sampled,
      min_annual_occupancy = min(annual_occupancy, na.rm = TRUE),
      max_annual_occupancy = max(annual_occupancy, na.rm = TRUE),
      .groups = "drop"
    )

  month_summary <-
    taxon_time_series |>
    group_by(OTU, month) |>
    summarize(
      monthly_occupancy = mean(active_observed, na.rm = TRUE),
      monthly_smoothed_activity = mean(active_smoothed, na.rm = TRUE),
      mean_abundance = mean(abundance, na.rm = TRUE),
      mean_smoothed_abundance = mean(abundance_smoothed, na.rm = TRUE),
      .groups = "drop"
    ) |>
    group_by(OTU) |>
    mutate(
      max_monthly_activity = max(monthly_smoothed_activity, na.rm = TRUE),
      max_monthly_abundance = max(mean_smoothed_abundance, na.rm = TRUE),
      peak_month = mean_smoothed_abundance >= max_monthly_abundance * seasonal_recurrence_cutoff
    ) |>
    summarize(
      n_peak_months = sum(peak_month),
      peak_window_active_samples = sum(monthly_occupancy[peak_month] * n(), na.rm = TRUE),
      peak_months = paste(month[peak_month], collapse = ","),
      max_monthly_activity = max(monthly_smoothed_activity, na.rm = TRUE),
      median_monthly_activity = median(monthly_smoothed_activity, na.rm = TRUE),
      min_monthly_activity = min(monthly_smoothed_activity, na.rm = TRUE),
      seasonal_amplitude = max_monthly_activity - median_monthly_activity,
      seasonal_relative_amplitude = seasonal_amplitude / (max_monthly_activity + 1e-12),
      seasonal_peak_to_background = max_monthly_activity / (median_monthly_activity + 1e-12),
      max_monthly_abundance = max(mean_smoothed_abundance, na.rm = TRUE),
      median_monthly_abundance = median(mean_smoothed_abundance, na.rm = TRUE),
      seasonal_abundance_amplitude = max_monthly_abundance - median_monthly_abundance,
      seasonal_abundance_relative_amplitude = seasonal_abundance_amplitude / (max_monthly_abundance + 1e-12),
      seasonal_abundance_peak_to_background = max_monthly_abundance / (median_monthly_abundance + 1e-12),
      .groups = "drop"
    )

  circular_summary <-
    taxon_time_series |>
    filter(active_observed) |>
    group_by(OTU) |>
    summarize(
      seasonal_x = weighted.mean(cos(theta), w = pmax(abundance_smoothed, 0), na.rm = TRUE),
      seasonal_y = weighted.mean(sin(theta), w = pmax(abundance_smoothed, 0), na.rm = TRUE),
      seasonal_concentration = sqrt(seasonal_x^2 + seasonal_y^2),
      seasonal_peak_doy = ((atan2(seasonal_y, seasonal_x) %% (2 * pi)) / (2 * pi)) * 365.25 + 1,
      .groups = "drop"
    )

  annual_peak_summary <-
    taxon_time_series |>
    filter(active_observed) |>
    group_by(OTU, year) |>
    mutate(annual_total_abundance = sum(abundance_smoothed, na.rm = TRUE)) |>
    slice_max(order_by = abundance_smoothed, n = 1, with_ties = FALSE) |>
    ungroup() |>
    group_by(OTU) |>
    summarize(
      annual_peak_years = n(),
      annual_peak_x = mean(cos(theta), na.rm = TRUE),
      annual_peak_y = mean(sin(theta), na.rm = TRUE),
      annual_peak_doy_concentration = sqrt(annual_peak_x^2 + annual_peak_y^2),
      annual_peak_doy = ((atan2(annual_peak_y, annual_peak_x) %% (2 * pi)) / (2 * pi)) * 365.25 + 1,
      max_single_year_peak_fraction = max(annual_total_abundance, na.rm = TRUE) / sum(annual_total_abundance, na.rm = TRUE),
      .groups = "drop"
    )

  harmonic_summary <-
    taxon_time_series |>
    group_by(OTU) |>
    group_modify(~ {
      y <- .x$abundance_smoothed
      y_scaled <- y / (max(y, na.rm = TRUE) + 1e-12)

      if (sum(is.finite(y_scaled)) < 10 || max(y_scaled, na.rm = TRUE) <= 0) {
        return(tibble(
          seasonal_harmonic_r2 = 0,
          seasonal_harmonic_amplitude = 0
        ))
      }

      fit <- lm(y_scaled ~ sin(theta) + cos(theta), data = .x)
      fit_summary <- summary(fit)
      coefs <- coef(fit)

      tibble(
        seasonal_harmonic_r2 = fit_summary$r.squared,
        seasonal_harmonic_amplitude = sqrt(coefs[["sin(theta)"]]^2 + coefs[["cos(theta)"]]^2)
      )
    }) |>
    ungroup()

  activity_summary <-
    taxon_time_series |>
    group_by(OTU) |>
    summarize(
      occupancy = mean(active_observed, na.rm = TRUE),
      presence_frequency = mean(active_observed, na.rm = TRUE),
      active_sample_count = sum(active_observed, na.rm = TRUE),
      smoothed_occupancy = mean(active_smoothed, na.rm = TRUE),
      mean_abundance_when_active = mean(abundance[active_observed], na.rm = TRUE),
      .groups = "drop"
    ) |>
    left_join(run_summary, by = "OTU") |>
    left_join(annual_summary, by = "OTU") |>
    left_join(month_summary, by = "OTU") |>
    left_join(circular_summary, by = "OTU") |>
    left_join(annual_peak_summary, by = "OTU") |>
    left_join(harmonic_summary, by = "OTU") |>
    mutate(
      max_active_run = replace_na(max_active_run, 0),
      mean_active_run = replace_na(mean_active_run, 0),
      seasonal_concentration = replace_na(seasonal_concentration, 0),
      annual_peak_doy_concentration = replace_na(annual_peak_doy_concentration, 0),
      annual_peak_years = replace_na(annual_peak_years, 0),
      max_single_year_peak_fraction = replace_na(max_single_year_peak_fraction, 1),
      seasonal_harmonic_r2 = replace_na(seasonal_harmonic_r2, 0),
      seasonal_harmonic_amplitude = replace_na(seasonal_harmonic_amplitude, 0),
      persistent = occupancy >= persistent_cutoff,
      candidate_seasonal =
        !persistent &
        years_detected >= seasonal_min_years &
        annual_recurrence >= seasonal_recurrence_cutoff &
        annual_peak_years >= seasonal_min_peak_years &
        max_single_year_peak_fraction <= seasonal_max_single_year_peak_fraction &
        n_peak_months >= seasonal_min_peak_months &
        n_peak_months <= seasonal_max_peak_months &
        (
          seasonal_concentration >= seasonal_concentration_cutoff |
          seasonal_peak_to_background >= seasonal_peak_to_background_cutoff |
          seasonal_relative_amplitude >= seasonal_relative_amplitude_cutoff |
          seasonal_abundance_peak_to_background >= seasonal_abundance_peak_to_background_cutoff |
          seasonal_abundance_relative_amplitude >= seasonal_abundance_relative_amplitude_cutoff |
          annual_peak_doy_concentration >= seasonal_peak_doy_concentration_cutoff |
          (
            seasonal_harmonic_r2 >= seasonal_harmonic_r2_cutoff &
            seasonal_harmonic_amplitude >= seasonal_harmonic_amplitude_cutoff
          )
        ),
      seasonal_temporal_support =
        presence_frequency >= seasonal_min_presence_frequency &
        (
          max_active_run >= seasonal_min_max_run |
          peak_window_active_samples >= seasonal_min_peak_window_active_samples
        ),
      seasonal = candidate_seasonal & seasonal_temporal_support,
      episodic_transient =
        !persistent &
        !seasonal &
        max_single_year_peak_fraction >= transient_single_year_dominance_cutoff &
        seasonal_abundance_peak_to_background >= transient_peak_to_background_cutoff,
      transient =
        !persistent &
        !seasonal &
        (
          occupancy <= transient_cutoff |
          episodic_transient |
          (candidate_seasonal & !seasonal_temporal_support)
        ),
      activity_pattern = case_when(
        persistent ~ "persistent",
        seasonal ~ "seasonal",
        transient ~ "transient",
        TRUE ~ "unpatterned"
      )
    ) |>
    select(
      OTU,
      activity_pattern,
      occupancy,
      presence_frequency,
      active_sample_count,
      smoothed_occupancy,
      annual_recurrence,
      years_detected,
      max_active_run,
      mean_active_run,
      n_peak_months,
      peak_window_active_samples,
      peak_months,
      seasonal_concentration,
      seasonal_peak_doy,
      seasonal_amplitude,
      seasonal_relative_amplitude,
      seasonal_peak_to_background,
      seasonal_abundance_amplitude,
      seasonal_abundance_relative_amplitude,
      seasonal_abundance_peak_to_background,
      annual_peak_years,
      annual_peak_doy_concentration,
      annual_peak_doy,
      max_single_year_peak_fraction,
      seasonal_harmonic_r2,
      seasonal_harmonic_amplitude,
      episodic_transient,
      mean_abundance_when_active
    )

  activity_summary
}
```

# Import and prepare data

```{r import_data}
sample_design <- read.delim(paths$design, header = TRUE, row.names = 1)

otu_counts <- read.otu(shared = paths$shared, cutoff = "0.03")
otu_taxonomy <- read.tax(taxonomy = paths$taxonomy, format = "rdp")

otu_counts <- otu_counts[rownames(otu_counts) != "RGMockComm", ]

set.seed(12742899)
otu_counts <- rrarefy(otu_counts, min(rowSums(otu_counts)))
otu_rel_abund <- make_relative_abundance(otu_counts)
```

# Prepare time-series data

```{r prepare_timeseries}
timeseries_counts <- otu_counts[str_which(rownames(otu_counts), "UL"), ]

timeseries_design <- read_csv(paths$timeseries_design)
timeseries_counts <- timeseries_counts[match(timeseries_design$sample.name, rownames(timeseries_counts)), ]

timeseries_rna_counts <- timeseries_counts[timeseries_design$sample.type == "RNA", ]
timeseries_dna_counts_raw <- timeseries_counts[timeseries_design$sample.type == "DNA", ]

# If an OTU is detected in RNA, it must also have been present in the total
# DNA pool. Add one pseudo-read to the DNA count table wherever RNA is detected.
timeseries_rna_detected <- make_presence(timeseries_rna_counts)
timeseries_dna_counts <- timeseries_dna_counts_raw + timeseries_rna_detected

timeseries_rna_rel_abund <- make_relative_abundance(timeseries_rna_counts)
timeseries_dna_rel_abund <- make_relative_abundance(timeseries_dna_counts)

timeseries_dna_presence <- make_presence(timeseries_dna_counts)
timeseries_rna_presence <- make_presence(timeseries_rna_counts)

env_data <- read.table(paths$env_data, sep = "\t", header = TRUE)
env_data$date <- as.Date(parse_date_time(env_data$date, "m d y"))
env_data$doc[env_data$doc == "**"] <- NA
env_data$doc <- as.numeric(env_data$doc)

timeseries_design <- left_join(
  timeseries_design,
  env_data |> select(sample.id, date),
  by = "sample.id"
)

env_data <- env_data[env_data$date %in% timeseries_design$date, ]

timeseries_long <- make_long_otu_table(
  otu_table = timeseries_rna_rel_abund,
  metadata = env_data,
  value_name = "relative_abundance"
)

timeseries_presence_long <- make_long_otu_table(
  otu_table = timeseries_rna_presence,
  metadata = env_data,
  value_name = "presence"
)
```

# Identify soil-associated taxa

```{r identify_soil_taxa}
soil_counts <- otu_counts[str_which(rownames(otu_counts), "RGSoil"), ]
soil_counts_detected <- soil_counts[, colSums(soil_counts) > 0]

soil_taxa <- colnames(soil_counts_detected)
nonsoil_taxa <- colnames(soil_counts[, colSums(soil_counts) == 0])

stopifnot(length(soil_taxa) + length(nonsoil_taxa) == ncol(otu_counts))

soil_max_rel_abund <- apply(make_relative_abundance(soil_counts), MARGIN = 2, max)
soil_common_taxa <- names(soil_max_rel_abund[soil_max_rel_abund >= 0.001])
soil_rare_taxa <- names(soil_max_rel_abund[soil_max_rel_abund < 0.001])

stopifnot(length(soil_common_taxa) + length(soil_rare_taxa) == ncol(otu_counts))
```

# Prepare transect data

```{r prepare_transect}
transect_counts <- otu_counts[str_which(rownames(otu_counts), "RG"), ]

transect_rna_counts <- transect_counts[str_which(rownames(transect_counts), "RGc"), ]
transect_dna_counts <- transect_counts[str_which(rownames(transect_counts), "RGD"), ]

transect_dna_presence <- make_presence(transect_dna_counts + transect_rna_counts)
transect_rna_presence <- make_presence(transect_rna_counts)
transect_dna_only_presence <- make_presence(transect_dna_counts)

taxa_active_in_transect <- colnames(transect_rna_presence[, colSums(transect_rna_presence) > 0])
taxa_present_in_transect <- colnames(transect_dna_presence[, colSums(transect_dna_presence) > 0])

soil_shared_rna_taxa <- intersect(soil_taxa, taxa_active_in_transect)
soil_shared_dna_taxa <- intersect(soil_taxa, taxa_present_in_transect)
```

# Define active taxa in the time series

```{r active_timeseries_taxa}
timeseries_rna_active_taxa <-
  timeseries_rna_presence[, colSums(timeseries_rna_presence) > 0, drop = FALSE]

# Soil-derived taxa detected in the transect DNA pool and active at least once
# in the time-series RNA pool.
timeseries_rna_soil_shared_dna_presence <-
  timeseries_rna_active_taxa[, colnames(timeseries_rna_active_taxa) %in% soil_shared_dna_taxa, drop = FALSE]

# Soil-derived taxa active in the transect RNA pool and active at least once
# in the time-series RNA pool.
timeseries_rna_soil_shared_rna_presence <-
  timeseries_rna_active_taxa[, colnames(timeseries_rna_active_taxa) %in% soil_shared_rna_taxa, drop = FALSE]

soil_shared_dna_active_taxa <-
  tibble(OTU = colnames(timeseries_rna_soil_shared_dna_presence))

soil_shared_rna_active_taxa <-
  tibble(OTU = colnames(timeseries_rna_soil_shared_rna_presence))

stopifnot(exists("timeseries_rna_soil_shared_dna_presence"))
stopifnot(exists("timeseries_rna_soil_shared_rna_presence"))
```

# Classify soil-associated activity patterns

```{r classify_soil_activity_patterns}
soil_activity_classes <- classify_activity_patterns(
  long_data = timeseries_long,
  presence_table = timeseries_rna_soil_shared_dna_presence,
  taxa = colnames(timeseries_rna_soil_shared_dna_presence),
  persistent_cutoff = 0.90,
  transient_cutoff = 0.10,
  seasonal_min_years = 2,
  seasonal_min_peak_months = 2,
  seasonal_max_peak_months = 4,
  seasonal_concentration_cutoff = 0.50,
  seasonal_recurrence_cutoff = 0.50,
  seasonal_min_presence_frequency = 0.05,
  seasonal_min_max_run = 2,
  seasonal_min_peak_window_active_samples = 4,
  seasonal_peak_to_background_cutoff = 3,
  seasonal_relative_amplitude_cutoff = 0.50,
  seasonal_abundance_peak_to_background_cutoff = 3,
  seasonal_abundance_relative_amplitude_cutoff = 0.50,
  seasonal_peak_doy_concentration_cutoff = 0.60,
  seasonal_min_peak_years = 2,
  seasonal_max_single_year_peak_fraction = 0.75,
  transient_single_year_dominance_cutoff = 0.75,
  transient_peak_to_background_cutoff = 3,
  seasonal_harmonic_r2_cutoff = 0.10,
  seasonal_harmonic_amplitude_cutoff = 0.25,
  smooth_activity = TRUE,
  smooth_window = 3
)

soil_activity_classes_unsmoothed <- classify_activity_patterns(
  long_data = timeseries_long,
  presence_table = timeseries_rna_soil_shared_dna_presence,
  taxa = colnames(timeseries_rna_soil_shared_dna_presence),
  persistent_cutoff = 0.90,
  transient_cutoff = 0.10,
  seasonal_min_years = 2,
  seasonal_min_peak_months = 2,
  seasonal_max_peak_months = 4,
  seasonal_concentration_cutoff = 0.50,
  seasonal_recurrence_cutoff = 0.50,
  seasonal_min_presence_frequency = 0.05,
  seasonal_min_max_run = 2,
  seasonal_min_peak_window_active_samples = 4,
  seasonal_peak_to_background_cutoff = 3,
  seasonal_relative_amplitude_cutoff = 0.50,
  seasonal_abundance_peak_to_background_cutoff = 3,
  seasonal_abundance_relative_amplitude_cutoff = 0.50,
  seasonal_peak_doy_concentration_cutoff = 0.60,
  seasonal_min_peak_years = 2,
  seasonal_max_single_year_peak_fraction = 0.75,
  transient_single_year_dominance_cutoff = 0.75,
  transient_peak_to_background_cutoff = 3,
  seasonal_harmonic_r2_cutoff = 0.10,
  seasonal_harmonic_amplitude_cutoff = 0.25,
  smooth_activity = FALSE
)

soil_activity_classes |> count(activity_pattern)

soil_smoothing_sensitivity <-
  soil_activity_classes_unsmoothed |>
  select(OTU, activity_pattern_unsmoothed = activity_pattern) |>
  left_join(
    soil_activity_classes |>
      select(OTU, activity_pattern_smoothed = activity_pattern),
    by = "OTU"
  )

soil_smoothing_sensitivity |> count(activity_pattern_unsmoothed, activity_pattern_smoothed)

soil_persistent_taxa <-
  soil_activity_classes |>
  filter(activity_pattern == "persistent")

soil_seasonal_taxa <-
  soil_activity_classes |>
  filter(activity_pattern == "seasonal")

soil_transient_taxa <-
  soil_activity_classes |>
  filter(activity_pattern == "transient")

soil_unpatterned_taxa <-
  soil_activity_classes |>
  filter(activity_pattern == "unpatterned")
```

## Visualize classifier-derived soil activity groups

```{r visualize_soil_activity_groups}
plot_relative_abundance_facets(timeseries_long, soil_persistent_taxa$OTU, x_var = date)
plot_relative_abundance_facets(timeseries_long, soil_seasonal_taxa$OTU, x_var = date)
plot_relative_abundance_facets(timeseries_long, soil_transient_taxa$OTU[1:min(200, nrow(soil_transient_taxa))], x_var = date)
plot_relative_abundance_facets(timeseries_long, soil_unpatterned_taxa$OTU[1:min(200, nrow(soil_unpatterned_taxa))], x_var = date)
```

# Contribution of classifier-derived soil-associated taxa to active community

```{r soil_contribution_by_activity}
soil_persistent_abundance <- summarize_taxon_abundance(timeseries_long, soil_persistent_taxa$OTU, env_data)
soil_seasonal_abundance <- summarize_taxon_abundance(timeseries_long, soil_seasonal_taxa$OTU, env_data)
soil_transient_abundance <- summarize_taxon_abundance(timeseries_long, soil_transient_taxa$OTU, env_data)

bind_rows(
  soil_persistent_abundance |> mutate(activity_pattern = "persistent"),
  soil_seasonal_abundance |> mutate(activity_pattern = "seasonal"),
  soil_transient_abundance |> mutate(activity_pattern = "transient")
) |>
  ggplot(aes(x = date, y = relative_abundance * 100, color = activity_pattern)) +
  geom_line(alpha = 0.9, linewidth = 0.8) +
  labs(x = "", y = "Percent abundance in lake\nof shared soil taxa") +
  scale_color_manual(values = activity_colors, drop = FALSE) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

ggsave(
  filename = "soil_taxa_percent_contribution_all.pdf",
  path = paths$figures,
  width = 6,
  height = 4.5
)
```

# Classify aquatic-only taxa

```{r classify_aquatic_activity_patterns}
timeseries_aquatic_rna_presence <-
  timeseries_rna_active_taxa[, !(colnames(timeseries_rna_active_taxa) %in% soil_shared_dna_taxa), drop = FALSE]

aquatic_only_taxa <- tibble(OTU = colnames(timeseries_aquatic_rna_presence))

aquatic_activity_classes <- classify_activity_patterns(
  long_data = timeseries_long,
  presence_table = timeseries_aquatic_rna_presence,
  taxa = colnames(timeseries_aquatic_rna_presence),
  persistent_cutoff = 0.90,
  transient_cutoff = 0.10,
  seasonal_min_years = 2,
  seasonal_min_peak_months = 2,
  seasonal_max_peak_months = 4,
  seasonal_concentration_cutoff = 0.50,
  seasonal_recurrence_cutoff = 0.50,
  seasonal_min_presence_frequency = 0.05,
  seasonal_min_max_run = 2,
  seasonal_min_peak_window_active_samples = 4,
  seasonal_peak_to_background_cutoff = 3,
  seasonal_relative_amplitude_cutoff = 0.50,
  seasonal_abundance_peak_to_background_cutoff = 3,
  seasonal_abundance_relative_amplitude_cutoff = 0.50,
  seasonal_peak_doy_concentration_cutoff = 0.60,
  seasonal_min_peak_years = 2,
  seasonal_max_single_year_peak_fraction = 0.75,
  transient_single_year_dominance_cutoff = 0.75,
  transient_peak_to_background_cutoff = 3,
  seasonal_harmonic_r2_cutoff = 0.10,
  seasonal_harmonic_amplitude_cutoff = 0.25,
  smooth_activity = TRUE,
  smooth_window = 3
)

aquatic_activity_classes_unsmoothed <- classify_activity_patterns(
  long_data = timeseries_long,
  presence_table = timeseries_aquatic_rna_presence,
  taxa = colnames(timeseries_aquatic_rna_presence),
  persistent_cutoff = 0.90,
  transient_cutoff = 0.10,
  seasonal_min_years = 2,
  seasonal_min_peak_months = 2,
  seasonal_max_peak_months = 4,
  seasonal_concentration_cutoff = 0.50,
  seasonal_recurrence_cutoff = 0.50,
  seasonal_min_presence_frequency = 0.05,
  seasonal_min_max_run = 2,
  seasonal_min_peak_window_active_samples = 4,
  seasonal_peak_to_background_cutoff = 3,
  seasonal_relative_amplitude_cutoff = 0.50,
  seasonal_abundance_peak_to_background_cutoff = 3,
  seasonal_abundance_relative_amplitude_cutoff = 0.50,
  seasonal_peak_doy_concentration_cutoff = 0.60,
  seasonal_min_peak_years = 2,
  seasonal_max_single_year_peak_fraction = 0.75,
  transient_single_year_dominance_cutoff = 0.75,
  transient_peak_to_background_cutoff = 3,
  seasonal_harmonic_r2_cutoff = 0.10,
  seasonal_harmonic_amplitude_cutoff = 0.25,
  smooth_activity = FALSE
)

aquatic_activity_classes |> count(activity_pattern)

aquatic_smoothing_sensitivity <-
  aquatic_activity_classes_unsmoothed |>
  select(OTU, activity_pattern_unsmoothed = activity_pattern) |>
  left_join(
    aquatic_activity_classes |>
      select(OTU, activity_pattern_smoothed = activity_pattern),
    by = "OTU"
  )

aquatic_smoothing_sensitivity |> count(activity_pattern_unsmoothed, activity_pattern_smoothed)

aquatic_persistent_taxa <-
  aquatic_activity_classes |>
  filter(activity_pattern == "persistent")

aquatic_seasonal_taxa <-
  aquatic_activity_classes |>
  filter(activity_pattern == "seasonal")

aquatic_transient_taxa <-
  aquatic_activity_classes |>
  filter(activity_pattern == "transient")

aquatic_unpatterned_taxa <-
  aquatic_activity_classes |>
  filter(activity_pattern == "unpatterned")
```

## Visualize classifier-derived aquatic activity groups

```{r visualize_aquatic_activity_groups}
plot_relative_abundance_facets(timeseries_long, aquatic_persistent_taxa$OTU, x_var = date)
plot_relative_abundance_facets(timeseries_long, aquatic_seasonal_taxa$OTU[1:min(200, nrow(aquatic_seasonal_taxa))], x_var = date)
plot_relative_abundance_facets(timeseries_long, aquatic_transient_taxa$OTU[1:min(200, nrow(aquatic_transient_taxa))], x_var = date)
plot_relative_abundance_facets(timeseries_long, aquatic_unpatterned_taxa$OTU[1:min(200, nrow(aquatic_unpatterned_taxa))], x_var = date)
```

# Contribution of classifier-derived aquatic-only taxa to active community

```{r aquatic_contribution_by_activity}
aquatic_persistent_abundance <- summarize_taxon_abundance(timeseries_long, aquatic_persistent_taxa$OTU, env_data)
aquatic_seasonal_abundance <- summarize_taxon_abundance(timeseries_long, aquatic_seasonal_taxa$OTU, env_data)
aquatic_transient_abundance <- summarize_taxon_abundance(timeseries_long, aquatic_transient_taxa$OTU, env_data)

bind_rows(
  aquatic_persistent_abundance |> mutate(activity_pattern = "persistent"),
  aquatic_seasonal_abundance |> mutate(activity_pattern = "seasonal"),
  aquatic_transient_abundance |> mutate(activity_pattern = "transient")
) |>
  ggplot(aes(x = date, y = relative_abundance, color = activity_pattern)) +
  geom_line(alpha = 0.9, linewidth = 0.8) +
  scale_color_manual(values = activity_colors, drop = FALSE) +
  labs(x = "Date", y = "Relative abundance") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Rare/common status in the time series

```{r classify_time_series_abundance_type}
timeseries_long <-
  timeseries_long |>
  mutate(abundance_type = if_else(relative_abundance >= 0.001, "common", "rare"))
```

# Contribution of all shared taxa to active community

```{r all_shared_taxa_contribution}
all_shared_taxa <- soil_shared_dna_active_taxa$OTU

total_shared_abundance <-
  timeseries_long |>
  filter(OTU %in% all_shared_taxa) |>
  group_by(sample_id) |>
  summarize(aggregate_abundance = sum(relative_abundance), .groups = "drop")

rare_shared_abundance <-
  timeseries_long |>
  filter(OTU %in% all_shared_taxa, abundance_type == "rare") |>
  group_by(sample_id) |>
  summarize(rare_abundance = sum(relative_abundance), .groups = "drop")

common_shared_abundance <-
  timeseries_long |>
  filter(OTU %in% all_shared_taxa, abundance_type == "common") |>
  group_by(sample_id) |>
  summarize(common_abundance = sum(relative_abundance), .groups = "drop")

shared_abundance_summary <-
  total_shared_abundance |>
  left_join(rare_shared_abundance, by = "sample_id") |>
  left_join(common_shared_abundance, by = "sample_id") |>
  mutate(across(contains("abundance"), ~ replace_na(.x, 0))) |>
  pivot_longer(
    cols = contains("abundance"),
    names_to = "abundance_type",
    values_to = "abundance"
  ) |>
  left_join(env_data |> select(sample.id, date), by = c("sample_id" = "sample.id"))

shared_abundance_summary |>
  ggplot(aes(x = date, y = abundance * 100, color = abundance_type)) +
  geom_line(alpha = 0.9, linewidth = 0.8) +
  facet_wrap(
    ~abundance_type,
    labeller = labeller(
      abundance_type = c(
        aggregate_abundance = "All Taxa",
        common_abundance = "Common Taxa",
        rare_abundance = "Rare Taxa"
      )
    )
  ) +
  guides(color = "none", linetype = "none") +
  labs(
    x = "",
    y = "Percent abundance
from active shared taxa"
  ) +
  scale_color_manual(values = abundance_colors, drop = FALSE) +
  #scale_linetype_manual(values = abundance_linetypes, drop = FALSE) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y")  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  filename = "percent_contribution_of_active_shared_taxa_in_aquatic_community.pdf",
  path = paths$figures,
  height = 4,
  width = 12,
  units = "in"
)
```

# Percent of shared taxa that are common when present

```{r percent_common_when_present}
timeseries_presence_lookup <-
  timeseries_presence_long |>
  select(sample_id, OTU, presence)

shared_common_when_present <-
  timeseries_long |>
  left_join(timeseries_presence_lookup, by = c("sample_id", "OTU")) |>
  filter(OTU %in% all_shared_taxa, presence == 1) |>
  group_by(sample_id) |>
  summarize(
    total_present = n(),
    total_common = sum(abundance_type == "common"),
    total_rare = sum(abundance_type == "rare"),
    percent_common = total_common / total_present * 100,
    percent_rare = total_rare / total_present * 100,
    .groups = "drop"
  ) |>
  mutate(mean_percent_common = mean(percent_common, na.rm = TRUE))

shared_common_when_present |>
  left_join(env_data |> select(sample.id, date), by = c("sample_id" = "sample.id")) |>
  ggplot(aes(x = date, y = percent_common)) +
  geom_line(alpha = 0.6, linewidth = 0.7) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(x = "Date", y = "Percent common among present shared taxa") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  "percent_of_shared_taxa_common_when_present.pdf",
  path = paths$figures,
  width = 6,
  height = 4,
  units = "in"
)
```

# Percent of time each shared OTU is common when present

```{r percent_time_common_when_present}
shared_percent_time_common <-
  timeseries_long |>
  left_join(timeseries_presence_lookup, by = c("sample_id", "OTU")) |>
  filter(OTU %in% all_shared_taxa, presence == 1) |>
  group_by(OTU) |>
  summarize(
    total_present = n(),
    total_common = sum(abundance_type == "common"),
    total_rare = sum(abundance_type == "rare"),
    percent_common = total_common / total_present * 100,
    percent_rare = total_rare / total_present * 100,
    .groups = "drop"
  ) |>
  left_join(soil_activity_classes, by = "OTU")

shared_percent_time_common |>
  filter(activity_pattern != "unpatterned") |>
  mutate(activity_pattern = factor(activity_pattern, levels = c("transient", "seasonal", "persistent"))) |>
  ggplot(aes(x = activity_pattern, y = percent_common, color = activity_pattern)) +
  geom_boxplot() +
  guides(color = "none") +
  labs(x = "Activity pattern", y = "Percent of time
common when present") +
  scale_color_manual(values = activity_colors, drop = FALSE)

ggsave(
  "shared_taxa_percent_of_times_common_when_present.pdf",
  path = paths$figures,
  width = 6,
  height = 4,
  units = "in"
)
```

# Example OTUs representing activity patterns

```{r example_otus}
timeseries_dna_long <- make_timeseries_long(
  otu_table = timeseries_dna_rel_abund,
  metadata = env_data,
  molecule_type = "DNA"
)

timeseries_rna_long <- make_timeseries_long(
  otu_table = timeseries_rna_rel_abund,
  metadata = env_data,
  molecule_type = "RNA"
)

example_otus <- tribble(
  ~OTU, ~activity_pattern,
  "Otu00001", "persistent",
  "Otu00078", "seasonal",
  "Otu00042", "transient"
)

plot_example_dna_rna_trajectories(
  dna_long = timeseries_dna_long,
  rna_long = timeseries_rna_long,
  example_taxa = example_otus
)

ggsave(
  filename = "example_dna_rna_activity_trajectories.pdf",
  path = paths$figures,
  width = 7,
  height = 6,
  units = "in"
)
```


# Compare maximum relative abundance across soil, transect, and time series

```{r maximum_relative_abundance_comparison}
soil_taxa_relative_abundance <-
  soil_counts_detected |>
  as.data.frame() |>
  make_relative_abundance() |>
  rownames_to_column(var = "sample_id") |>
  pivot_longer(
    cols = -sample_id,
    names_to = "OTU",
    values_to = "soil_relative_abundance"
  ) |>
  mutate(soil_abundance_type = if_else(soil_relative_abundance > 0.001, "common", "rare"))

soil_taxa_max_relative_abundance <-
  soil_taxa_relative_abundance |>
  group_by(OTU) |>
  summarize(
    soil_max_relative_abundance = max(soil_relative_abundance, na.rm = TRUE),
    soil_max_abundance_type = if_else(soil_max_relative_abundance > 0.001, "common", "rare"),
    .groups = "drop"
  )

transect_dna_max_relative_abundance <- get_max_relative_abundance(transect_dna_counts, "transect_dna")
transect_rna_max_relative_abundance <- get_max_relative_abundance(transect_rna_counts, "transect_rna")
timeseries_dna_max_relative_abundance <- get_max_relative_abundance(timeseries_dna_counts, "timeseries_dna")
timeseries_rna_max_relative_abundance <- get_max_relative_abundance(timeseries_rna_counts, "timeseries_rna")

shared_taxa_active_in_timeseries <-
  timeseries_rna_active_taxa |>
  as.data.frame() |>
  select(any_of(colnames(soil_counts_detected))) |>
  colnames()

combined_activity_classes <-
  bind_rows(
    aquatic_activity_classes,
    soil_activity_classes
  ) |>
  distinct(OTU, .keep_all = TRUE)

shared_taxa_summary <-
  tibble(OTU = shared_taxa_active_in_timeseries) |>
  left_join(soil_taxa_max_relative_abundance, by = "OTU") |>
  left_join(transect_dna_max_relative_abundance, by = "OTU") |>
  left_join(transect_rna_max_relative_abundance, by = "OTU") |>
  left_join(timeseries_dna_max_relative_abundance, by = "OTU") |>
  left_join(timeseries_rna_max_relative_abundance, by = "OTU") |>
  left_join(combined_activity_classes, by = "OTU")
```

## Soil versus aquatic transect maximum abundance

```{r plot_soil_vs_transect_max_abundance}
shared_taxa_summary |>
  ggplot(aes(
    x = soil_max_relative_abundance,
    y = transect_dna_max_relative_abundance,
    color = activity_pattern
  )) +
  geom_hline(yintercept = 0.001, color = "grey60") +
  geom_vline(xintercept = 0.001, color = "grey60") +
  geom_point(alpha = 0.35, size = 2.5, show.legend = FALSE) +
  labs(
    x = "Max. relative abundance\nin soil",
    y = "Max. relative abundance\nin aquatic transect",
    title = "Presence in lake (DNA)"
  ) +
  #annotate("text", x = 10^(-1), y = 10^(-1), label = "DNA", size = 6, adj = 1) +
  scale_x_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  scale_y_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  scale_color_manual(values = activity_colors, drop = FALSE) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  facet_wrap(~activity_pattern, ncol = 2)

ggsave(filename = "shared_taxa_soil_vs_transect_dna.pdf", path = paths$figures, width = 8, height = 8, units = "in")

shared_taxa_summary |>
  ggplot(aes(
    x = soil_max_relative_abundance,
    y = transect_rna_max_relative_abundance,
    color = activity_pattern
  )) +
  geom_hline(yintercept = 0.001, color = "grey60") +
  geom_vline(xintercept = 0.001, color = "grey60") +
  geom_point(alpha = 0.35, size = 2.5, show.legend = FALSE) +
  labs(
    x = "Max. relative abundance\nin soil",
    y = "Max. relative abundance\nin aquatic transect",
    title = "Activity in lake (RNA)"
  ) +
  #annotate("text", x = 10^(-1), y = 10^(-1), label = "RNA", size = 6, adj = 1) +
  scale_x_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  scale_y_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  scale_color_manual(values = activity_colors, drop = FALSE) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  facet_wrap(~activity_pattern, ncol = 2)

ggsave(filename = "shared_taxa_soil_vs_transect_rna.pdf", path = paths$figures, width = 8, height = 8, units = "in")
```
