---
title: "Dynamics of aquatic and soil-derived bacteria in a temperate lake"
author: "Tajinder Singh, Nathan I. Wisnoski"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
header-includes:
- \usepackage{array}
- \usepackage{graphics}
---

# Overview

This analysis evaluates temporal dynamics of aquatic and soil-associated bacterial OTUs in a temperate lake. The workflow identifies taxa detected in soil, taxa shared between soil and aquatic samples, taxa unique to aquatic samples, and temporal activity patterns based on RNA detections through time.

# Setup

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

## Load packages and helper functions

```{r packages, results='hide'}
library(tidyverse)
library(vegan)
library(viridis)
library(lubridate)
library(readxl)
library(writexl)
library(colorspace)
library(ggpubr)

source("bin/mothur_tools.R")

se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(na.omit(x)))
}

cv <- function(x) {
  sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
}
```

## Define paths and constants

```{r paths-and-constants}
paths <- list(
  design = "data/UL.design.txt",
  shared = "data/ul_resgrad.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.opti_mcc.shared",
  taxonomy = "data/ul_resgrad.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.opti_mcc.0.03.cons.taxonomy",
  timeseries_design = "data/UL_timeseries_design.csv",
  env_data = "data/ul-seedbank.env.txt",
  soil_activity_manual = "soil_otu_activity_pattern/new.sortedotus.xlsx",
  aquatic_persistent_manual = "aquatic_otu_activity_pattern/new.persistent.classified.timeseries.aquatic.otus.rna.present.atl.once.csv",
  aquatic_seasonal_manual = "aquatic_otu_activity_pattern/new.seasonal.classified.timeseries.aquatic.otus.rna.present.atl.once.csv",
  aquatic_transient_manual = "aquatic_otu_activity_pattern/new.transient.classified.timeseries.aquatic.otus.rna.present.atl.once.csv",
  figures = "figures"
)

rare_common_threshold <- 0.001
persistent_threshold <- 0.90
transient_threshold <- 0.10
rarefaction_seed <- 12742899

dir.create(paths$figures, showWarnings = FALSE, recursive = TRUE)
```

## Figure theme

```{r figure-theme}
my_cols <- RColorBrewer::brewer.pal(n = 4, name = "Greys")[3:4]

activity_colors <- c(
  transient = "#A42F2F",
  seasonal = "#3A8639",
  persistent = "#3F2BBD",
  unpatterned = "grey50"
)

abundance_colors <- c(
  aggregate_abundance = "#460E5F",
  common_abundance = "#B16415",
  rare_abundance = "#237B75"
)

theme_set(
  theme_classic() +
    theme(
      axis.title = element_text(size = 14),
      axis.title.x = element_text(margin = margin(t = 15, b = 15)),
      axis.title.y = element_text(margin = margin(l = 15, r = 15)),
      axis.text = element_text(size = 12),
      axis.text.x = element_text(margin = margin(t = 5)),
      axis.text.y = element_text(margin = margin(r = 5)),
      axis.line.x = element_blank(),
      axis.line.y = element_blank(),
      axis.ticks.x = element_line(linewidth = 1),
      axis.ticks.y = element_line(linewidth = 1),
      axis.ticks.length = unit(0.1, "in"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1.5),
      legend.title = element_blank(),
      legend.text = element_text(size = 14),
      strip.text = element_text(size = 12),
      strip.background = element_blank()
    )
)
```

# Helper functions

```{r helper-functions}
make_rel_abund <- function(count_table) {
  decostand(count_table, method = "total")
}

make_presence <- function(count_table) {
  decostand(count_table, method = "pa")
}

make_long_otu_table <- function(otu_table, metadata, sample_col = "sample.id") {
  otu_table |>
    as_tibble() |>
    mutate(
      sample.id = metadata[[sample_col]],
      .before = 1
    ) |>
    mutate(sample.id = as.integer(sample.id)) |>
    pivot_longer(
      cols = -sample.id,
      names_to = "OTU",
      values_to = "relative_abundance"
    ) |>
    left_join(metadata, by = "sample.id")
}

summarize_taxon_abundance <- function(long_data, taxa, metadata) {
  long_data |>
    filter(OTU %in% taxa) |>
    group_by(sample.id) |>
    summarize(relative_abundance = sum(relative_abundance), .groups = "drop") |>
    left_join(metadata |> select(sample.id, date), by = "sample.id")
}

classify_activity_patterns <- function(presence_table,
                                       persistent_cutoff = persistent_threshold,
                                       transient_cutoff = transient_threshold) {
  timeseries_length <- nrow(presence_table)

  tibble(
    OTU = colnames(presence_table),
    occupancy = colSums(presence_table) / timeseries_length,
    persistent = occupancy > persistent_cutoff,
    transient = occupancy < transient_cutoff,
    activity_pattern = case_when(
      persistent ~ "persistent",
      transient ~ "transient",
      TRUE ~ "seasonal"
    )
  ) |>
    select(OTU, occupancy, activity_pattern)
}

classify_activity_patterns_with_seasonality <- function(long_data,
                                                        presence_table,
                                                        taxa,
                                                        persistent_cutoff = persistent_threshold,
                                                        transient_cutoff = transient_threshold,
                                                        seasonal_ratio_cutoff = 0.90) {
  timeseries_length <- nrow(presence_table)

  within_month_var <- long_data |>
    mutate(month = month(date)) |>
    filter(OTU %in% taxa) |>
    group_by(OTU, month) |>
    summarize(var_within_month = sd(relative_abundance), .groups = "drop") |>
    group_by(OTU) |>
    summarize(var_within_month_mean = mean(var_within_month, na.rm = TRUE), .groups = "drop")

  within_year_var <- long_data |>
    mutate(year = year(date)) |>
    filter(OTU %in% taxa) |>
    group_by(OTU, year) |>
    summarize(var_within_year = sd(relative_abundance), .groups = "drop") |>
    group_by(OTU) |>
    summarize(var_within_year_mean = mean(var_within_year, na.rm = TRUE), .groups = "drop")

  among_month_var <- long_data |>
    mutate(month = month(date)) |>
    filter(OTU %in% taxa) |>
    group_by(OTU, month) |>
    summarize(mean_abund = mean(relative_abundance), .groups = "drop") |>
    group_by(OTU) |>
    summarize(var_across_month = sd(mean_abund), .groups = "drop")

  activity_summary <- tibble(
    OTU = colnames(presence_table),
    occupancy = colSums(presence_table) / timeseries_length,
    persistent = occupancy > persistent_cutoff,
    transient = occupancy < transient_cutoff
  )

  activity_summary |>
    left_join(within_month_var, by = "OTU") |>
    left_join(within_year_var, by = "OTU") |>
    left_join(among_month_var, by = "OTU") |>
    mutate(
      seasonal_ratio = var_within_month_mean / var_within_year_mean,
      seasonal = !persistent & !transient & seasonal_ratio < seasonal_ratio_cutoff,
      activity_pattern = case_when(
        persistent ~ "persistent",
        transient ~ "transient",
        seasonal ~ "seasonal",
        TRUE ~ "unpatterned"
      )
    ) |>
    select(OTU, occupancy, seasonal_ratio, activity_pattern)
}

plot_activity_facets <- function(long_data, taxa, y_var = relative_abundance, x_var = date) {
  long_data |>
    filter(OTU %in% taxa) |>
    ggplot(aes(x = {{ x_var }}, y = {{ y_var }}, group = OTU)) +
    geom_line(alpha = 0.3) +
    facet_wrap(~OTU, scales = "free_y") +
    theme_minimal()
}

summarize_common_rare_abundance <- function(long_data, taxa, metadata) {
  total_abundance <- long_data |>
    filter(OTU %in% taxa) |>
    group_by(sample.id) |>
    summarize(aggregate_abundance = sum(relative_abundance), .groups = "drop")

  rare_abundance <- long_data |>
    filter(OTU %in% taxa, abundance_type == "rare") |>
    group_by(sample.id) |>
    summarize(rare_abundance = sum(relative_abundance), .groups = "drop")

  common_abundance <- long_data |>
    filter(OTU %in% taxa, abundance_type == "common") |>
    group_by(sample.id) |>
    summarize(common_abundance = sum(relative_abundance), .groups = "drop")

  total_abundance |>
    left_join(rare_abundance, by = "sample.id") |>
    left_join(common_abundance, by = "sample.id") |>
    mutate(across(contains("abundance"), ~replace_na(.x, 0))) |>
    left_join(metadata |> select(sample.id, date), by = "sample.id")
}

make_max_rel_abund_table <- function(count_table, value_name, type_name) {
  count_table |>
    as.data.frame() |>
    make_rel_abund() |>
    t() |>
    as.data.frame() |>
    filter(rowSums(pick(everything())) > 0) |>
    mutate(
      "{value_name}" := apply(pick(everything()), 1, max),
      "{type_name}" := if_else(.data[[value_name]] > rare_common_threshold, "common", "rare")
    ) |>
    rownames_to_column(var = "OTU") |>
    select(OTU, all_of(value_name), all_of(type_name))
}
```

# Import and prepare data

```{r import-data}
sample_design <- read.delim(paths$design, header = TRUE, row.names = 1)

otu_counts <- read.otu(shared = paths$shared, cutoff = "0.03")
otu_taxonomy <- read.tax(taxonomy = paths$taxonomy, format = "rdp")

otu_counts <- otu_counts[rownames(otu_counts) != "RGMockComm", ]

set.seed(rarefaction_seed)
otu_counts <- rrarefy(otu_counts, min(rowSums(otu_counts)))
otu_rel_abund <- make_rel_abund(otu_counts)

sum(otu_counts == 0)
```

# Time-series samples

```{r prepare-timeseries-data}
timeseries_counts <- otu_counts[str_which(rownames(otu_counts), "UL"), ]

timeseries_design <- read_csv(paths$timeseries_design, show_col_types = FALSE)
timeseries_counts <- timeseries_counts[match(timeseries_design$sample.name, rownames(timeseries_counts)), ]

timeseries_rna_counts <- timeseries_counts[timeseries_design$sample.type == "RNA", ]
timeseries_dna_counts <- timeseries_counts[timeseries_design$sample.type == "DNA", ]

timeseries_rna_rel_abund <- make_rel_abund(timeseries_rna_counts)
timeseries_dna_rel_abund <- make_rel_abund(timeseries_dna_counts)

timeseries_rna_presence <- make_presence(timeseries_rna_counts)
timeseries_dna_presence <- make_presence(timeseries_dna_counts + timeseries_rna_counts)

env_data <- read.table(paths$env_data, sep = "\t", header = TRUE)
env_data$date <- as.Date(parse_date_time(env_data$date, "m d y"))
env_data$doc[env_data$doc == "**"] <- NA
env_data$doc <- as.numeric(env_data$doc)

timeseries_design <- left_join(
  timeseries_design,
  env_data |> select(sample.id, date),
  by = "sample.id"
)

env_data <- env_data |> filter(date %in% timeseries_design$date)

timeseries_long <- make_long_otu_table(timeseries_rna_rel_abund, env_data)
timeseries_presence_long <- make_long_otu_table(timeseries_rna_presence, env_data)
```

# Soil taxa and shared taxa

```{r soil-and-shared-taxa}
soil_counts <- otu_counts[str_which(rownames(otu_counts), "RGSoil"), ]
soil_counts_detected <- soil_counts[, colSums(soil_counts) > 0]

soil_taxa <- colnames(soil_counts_detected)
nonsoil_taxa <- colnames(soil_counts[, colSums(soil_counts) == 0])

length(soil_taxa) + length(nonsoil_taxa) == ncol(otu_counts)

soil_max_rel_abund <- apply(make_rel_abund(soil_counts), 2, max)
soil_common_taxa <- names(soil_max_rel_abund[soil_max_rel_abund >= rare_common_threshold])
soil_rare_taxa <- names(soil_max_rel_abund[soil_max_rel_abund < rare_common_threshold])

length(soil_common_taxa) + length(soil_rare_taxa) == ncol(otu_counts)
```

# Transect samples

```{r prepare-transect-data}
transect_counts <- otu_counts[str_which(rownames(otu_counts), "RG"), ]
transect_rna_counts <- transect_counts[str_which(rownames(transect_counts), "RGc"), ]
transect_dna_counts <- transect_counts[str_which(rownames(transect_counts), "RGD"), ]

transect_rna_presence <- make_presence(transect_rna_counts)
transect_dna_presence <- make_presence(transect_dna_counts + transect_rna_counts)
transect_dna_only_presence <- make_presence(transect_dna_counts)

transect_rna_taxa <- colnames(transect_rna_presence[, colSums(transect_rna_presence) > 0])
transect_dna_taxa <- colnames(transect_dna_presence[, colSums(transect_dna_presence) > 0])

shared_rna_taxa <- intersect(soil_taxa, transect_rna_taxa)
shared_dna_taxa <- intersect(soil_taxa, transect_dna_taxa)
```

# Activity patterns of shared taxa

```{r classify-shared-taxa}
timeseries_rna_active_once <- timeseries_rna_presence[, colSums(timeseries_rna_presence) > 0]

shared_dna_taxa_active_timeseries <- timeseries_rna_active_once[, colnames(timeseries_rna_active_once) %in% shared_dna_taxa]
shared_rna_taxa_active_timeseries <- timeseries_rna_active_once[, colnames(timeseries_rna_active_once) %in% shared_rna_taxa]

shared_activity_classes_metric <- classify_activity_patterns_with_seasonality(
  long_data = timeseries_long,
  presence_table = shared_dna_taxa_active_timeseries,
  taxa = colnames(shared_rna_taxa_active_timeseries)
)

shared_activity_classes_metric |> count(activity_pattern)
```

## Load manually curated shared-taxon activity classes

```{r load-manual-shared-activity-classes}
shared_persistent_taxa <- read_excel(paths$soil_activity_manual, sheet = "persistent")
shared_seasonal_taxa <- read_excel(paths$soil_activity_manual, sheet = "seasonal")
shared_transient_taxa <- read_excel(paths$soil_activity_manual, sheet = "transient")

shared_activity_classes <- bind_rows(
  shared_persistent_taxa,
  shared_seasonal_taxa,
  shared_transient_taxa
) |>
  rename(activity_pattern = active)

shared_activity_classes |> count(activity_pattern)
```

## Visualize manually curated shared-taxon activity classes

```{r plot-shared-activity-classes}
plot_activity_facets(timeseries_presence_long, shared_persistent_taxa$OTU) +
  scale_y_continuous(limits = c(-0.5, 1.5))

plot_activity_facets(timeseries_long, shared_persistent_taxa$OTU)

plot_activity_facets(timeseries_presence_long, shared_seasonal_taxa$OTU) +
  scale_y_continuous(limits = c(-0.5, 1.5))

plot_activity_facets(timeseries_long, shared_seasonal_taxa$OTU)

plot_activity_facets(timeseries_presence_long, shared_transient_taxa$OTU[1:60]) +
  scale_y_continuous(limits = c(-0.5, 1.5))

plot_activity_facets(timeseries_long, shared_transient_taxa$OTU[1:100])
```

# Contribution of shared taxa to the active aquatic community

```{r shared-taxon-contributions}
shared_persistent_abund <- summarize_taxon_abundance(
  long_data = timeseries_long,
  taxa = shared_persistent_taxa$OTU,
  metadata = env_data
)

shared_seasonal_abund <- summarize_taxon_abundance(
  long_data = timeseries_long,
  taxa = shared_seasonal_taxa$OTU,
  metadata = env_data
)

shared_transient_abund <- summarize_taxon_abundance(
  long_data = timeseries_long,
  taxa = shared_transient_taxa$OTU,
  metadata = env_data
)

shared_contribution_long <- bind_rows(
  shared_persistent_abund |> mutate(activity_pattern = "persistent"),
  shared_seasonal_abund |> mutate(activity_pattern = "seasonal"),
  shared_transient_abund |> mutate(activity_pattern = "transient")
)

shared_contribution_long |>
  mutate(activity_pattern = factor(activity_pattern, levels = c("transient", "seasonal", "persistent"))) |>
  ggplot(aes(x = date, y = relative_abundance * 100, color = activity_pattern)) +
  geom_line(alpha = 0.8) +
  labs(x = "", y = "Percent Abundance") +
  scale_y_continuous(limits = c(0, 65)) +
  scale_color_discrete_qualitative(palette = "Dark3") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "inside",
    legend.position.inside = c(0.15, 0.84)
  )

ggsave(
  filename = "soil_taxa_percent_contribution_all.pdf",
  width = 6,
  height = 4.5,
  path = paths$figures
)
```

# Aquatic-only taxa

```{r aquatic-only-taxa}
aquatic_only_presence <- timeseries_rna_active_once[, !(colnames(timeseries_rna_active_once) %in% shared_dna_taxa)]
aquatic_only_taxa <- tibble(OTU = colnames(aquatic_only_presence))

aquatic_activity_classes_metric <- classify_activity_patterns(aquatic_only_presence)
aquatic_activity_classes_metric |> count(activity_pattern)

write_xlsx(
  aquatic_activity_classes_metric,
  "classified_timeseries_aquatic_otus_rna_present_at_least_once.xlsx"
)
```

## Load manually curated aquatic-only activity classes

```{r load-manual-aquatic-activity-classes}
aquatic_persistent_taxa <- read_csv(paths$aquatic_persistent_manual, show_col_types = FALSE)
aquatic_seasonal_taxa <- read_csv(paths$aquatic_seasonal_manual, show_col_types = FALSE)
aquatic_transient_taxa <- read_csv(paths$aquatic_transient_manual, show_col_types = FALSE)

aquatic_activity_classes <- bind_rows(
  aquatic_persistent_taxa,
  aquatic_seasonal_taxa,
  aquatic_transient_taxa
) |>
  rename(activity_pattern = active)

aquatic_activity_classes |> count(activity_pattern)
```

## Visualize aquatic-only activity classes

```{r plot-aquatic-activity-classes}
plot_activity_facets(timeseries_presence_long, aquatic_persistent_taxa$OTU) +
  scale_y_continuous(limits = c(-0.5, 1.5))

plot_activity_facets(timeseries_presence_long, aquatic_seasonal_taxa$OTU[1:30]) +
  scale_y_continuous(limits = c(-0.5, 1.5))

plot_activity_facets(timeseries_presence_long, aquatic_transient_taxa$OTU[1:30]) +
  scale_y_continuous(limits = c(-0.5, 1.5))
```

# Contribution of aquatic-only taxa to the active aquatic community

```{r aquatic-only-contributions}
aquatic_persistent_abund <- summarize_taxon_abundance(
  long_data = timeseries_long,
  taxa = aquatic_persistent_taxa$OTU,
  metadata = env_data
)

aquatic_seasonal_abund <- summarize_taxon_abundance(
  long_data = timeseries_long,
  taxa = aquatic_seasonal_taxa$OTU,
  metadata = env_data
)

aquatic_transient_abund <- summarize_taxon_abundance(
  long_data = timeseries_long,
  taxa = aquatic_transient_taxa$OTU,
  metadata = env_data
)

bind_rows(
  aquatic_persistent_abund |> mutate(activity_pattern = "persistent"),
  aquatic_seasonal_abund |> mutate(activity_pattern = "seasonal"),
  aquatic_transient_abund |> mutate(activity_pattern = "transient")
) |>
  mutate(activity_pattern = factor(activity_pattern, levels = c("transient", "seasonal", "persistent"))) |>
  ggplot(aes(x = date, y = relative_abundance, color = activity_pattern)) +
  geom_line(alpha = 0.8) +
  theme_minimal()
```

# Rare and common contributions among shared taxa

```{r rare-common-prep}
timeseries_long <- timeseries_long |>
  mutate(
    abundance_type = if_else(
      relative_abundance >= rare_common_threshold,
      "common",
      "rare"
    )
  )

timeseries_presence_long <- timeseries_presence_long |>
  rename(presence = relative_abundance)
```

## Partition shared-taxon abundance into rare and common components

```{r rare-common-shared-abundance}
shared_common_rare_wide <- summarize_common_rare_abundance(
  long_data = timeseries_long,
  taxa = shared_dna_taxa_active_timeseries |> colnames(),
  metadata = env_data
)

shared_common_rare_long <- shared_common_rare_wide |>
  pivot_longer(
    cols = contains("abundance"),
    names_to = "abundance_type",
    values_to = "abundance"
  )

shared_common_rare_long |>
  ggplot(aes(x = sample.id, y = abundance * 100, color = abundance_type)) +
  geom_line(alpha = 0.8) +
  facet_wrap(
    ~abundance_type,
    labeller = labeller(
      abundance_type = c(
        aggregate_abundance = "All Taxa",
        common_abundance = "Common Taxa",
        rare_abundance = "Rare Taxa"
      )
    )
  ) +
  guides(color = "none") +
  labs(
    x = "Weeks",
    y = "Percent Abundance\nfrom Active Shared Taxa"
  ) +
  scale_color_manual(values = abundance_colors) +
  theme(
    plot.margin = unit(c(1, 1, 1, 1), "cm"),
    aspect.ratio = 0.8,
    strip.text = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )

ggsave(
  filename = "percent_contribution_of_active_shared_taxa_in_aquatic_community.pdf",
  path = paths$figures,
  height = 10,
  width = 13,
  units = "in"
)
```

## Percent of present shared taxa that are common

```{r percent-common-when-present}
shared_presence_with_abundance_type <- timeseries_long |>
  select(sample.id, OTU, relative_abundance, abundance_type) |>
  left_join(
    timeseries_presence_long |> select(sample.id, OTU, presence),
    by = c("sample.id", "OTU")
  )

shared_percent_common_by_sample <- shared_presence_with_abundance_type |>
  filter(OTU %in% colnames(shared_dna_taxa_active_timeseries), presence == 1) |>
  group_by(sample.id) |>
  summarize(
    total_present = n(),
    total_common = sum(abundance_type == "common"),
    total_rare = sum(abundance_type == "rare"),
    percent_common = 100 * total_common / total_present,
    percent_rare = 100 * total_rare / total_present,
    .groups = "drop"
  )

shared_percent_common_by_sample |>
  ggplot(aes(x = sample.id, y = percent_common)) +
  geom_line(alpha = 0.3) +
  theme_minimal()
```

## Percent common by activity class

```{r percent-common-by-activity-class}
percent_common_by_activity_class <- function(long_data, presence_long, taxa, activity_pattern) {
  long_data |>
    select(sample.id, OTU, relative_abundance, abundance_type) |>
    left_join(
      presence_long |> select(sample.id, OTU, presence),
      by = c("sample.id", "OTU")
    ) |>
    filter(OTU %in% taxa, presence == 1) |>
    group_by(sample.id) |>
    summarize(
      total_present = n(),
      total_common = sum(abundance_type == "common"),
      total_rare = sum(abundance_type == "rare"),
      percent_common = 100 * total_common / total_present,
      percent_rare = 100 * total_rare / total_present,
      .groups = "drop"
    ) |>
    mutate(activity_pattern = activity_pattern)
}

shared_percent_common_by_activity <- bind_rows(
  percent_common_by_activity_class(timeseries_long, timeseries_presence_long, shared_persistent_taxa$OTU, "persistent"),
  percent_common_by_activity_class(timeseries_long, timeseries_presence_long, shared_seasonal_taxa$OTU, "seasonal"),
  percent_common_by_activity_class(timeseries_long, timeseries_presence_long, shared_transient_taxa$OTU, "transient")
)

shared_percent_common_by_activity |>
  mutate(activity_pattern = factor(activity_pattern, levels = c("transient", "seasonal", "persistent"))) |>
  ggplot(aes(x = activity_pattern, y = percent_common, color = activity_pattern)) +
  geom_boxplot() +
  guides(color = "none") +
  labs(
    x = "Activity Pattern",
    y = "Percent of taxa\ncommon when present"
  ) +
  theme(
    axis.title = element_text(face = "bold", size = 20),
    axis.text = element_text(size = 15),
    plot.margin = margin(20, 20),
    axis.title.y = element_text(hjust = 0.72)
  ) +
  scale_color_manual(values = activity_colors)

ggsave("percent_of_shared_taxa_common_when_present.pdf", 
       path = paths$figures,
       height = 4,
       width = 6,
       units = "in")
```

## Percent of time each shared taxon is common when present

```{r percent-time-common-when-present}
shared_percent_time_common <- shared_presence_with_abundance_type |>
  filter(OTU %in% colnames(shared_dna_taxa_active_timeseries), presence == 1) |>
  group_by(OTU) |>
  summarize(
    total_present = n(),
    total_common = sum(abundance_type == "common"),
    total_rare = sum(abundance_type == "rare"),
    percent_common = 100 * total_common / total_present,
    percent_rare = 100 * total_rare / total_present,
    .groups = "drop"
  ) |>
  left_join(shared_activity_classes, by = "OTU")

shared_percent_time_common |>
  mutate(activity_pattern = factor(activity_pattern, levels = c("transient", "seasonal", "persistent"))) |>
  ggplot(aes(x = activity_pattern, y = percent_common, color = activity_pattern)) +
  geom_boxplot() +
  guides(color = "none") +
  labs(
    x = "Activity Pattern",
    y = "Percent of time\ncommon when present"
  ) +
  theme(
    axis.title = element_text(face = "bold", size = 20),
    axis.text = element_text(size = 15),
    plot.margin = margin(20, 10),
    axis.title.y = element_text(hjust = 0.72)
  ) +
  scale_color_manual(values = activity_colors)

ggsave("shared_taxa_percent_of_times_common_when_present.pdf", 
       path = paths$figures,
       height = 4,
       width = 6,
       units = "in")
```

# Example OTUs representing activity patterns

```{r example-otus}
plot_example_otu <- function(presence_long, otu, color, output_file) {
  p <- presence_long |>
    filter(OTU == otu) |>
    ggplot(aes(x = date, y = presence)) +
    facet_wrap(~OTU) +
    geom_line(alpha = 1, color = color) +
    scale_y_continuous(limits = c(-0.5, 1.5)) +
    labs(x = "Weeks", y = "") +
    theme(
      axis.title = element_text(face = "bold", size = 20),
      axis.text.x = element_text(size = 12)
    )

  print(p)
  ggsave(output_file, plot = p, path = paths$figures)
}

plot_example_otu(timeseries_presence_long, "Otu00001", activity_colors[["persistent"]], "persistent_example_otu.pdf")
plot_example_otu(timeseries_presence_long, "Otu00078", activity_colors[["seasonal"]], "seasonal_example_otu.pdf")
plot_example_otu(timeseries_presence_long, "Otu00042", activity_colors[["transient"]], "transient_example_otu.pdf")
```

# Maximum relative abundance in soil, transect, and time series

```{r max-relative-abundance}
soil_taxa_rel_abund <- soil_counts_detected |>
  as.data.frame() |>
  make_rel_abund() |>
  rownames_to_column(var = "sample_id") |>
  pivot_longer(
    cols = -sample_id,
    names_to = "OTU",
    values_to = "soil_rel_abundance"
  ) |>
  mutate(soil_abundance_type = if_else(soil_rel_abundance > rare_common_threshold, "common", "rare"))

soil_taxa_max_rel_abund <- soil_taxa_rel_abund |>
  group_by(OTU) |>
  summarize(
    soil_max_rel_abundance = max(soil_rel_abundance),
    soil_max_abundance_type = if_else(soil_max_rel_abundance > rare_common_threshold, "common", "rare"),
    .groups = "drop"
  )

transect_dna_max_rel_abund <- make_max_rel_abund_table(
  transect_dna_counts,
  value_name = "transect_dna_max_rel_abundance",
  type_name = "transect_dna_max_abundance_type"
)

transect_rna_max_rel_abund <- make_max_rel_abund_table(
  transect_rna_counts,
  value_name = "transect_rna_max_rel_abundance",
  type_name = "transect_rna_max_abundance_type"
)

timeseries_dna_max_rel_abund <- make_max_rel_abund_table(
  timeseries_dna_counts,
  value_name = "timeseries_dna_max_rel_abundance",
  type_name = "timeseries_dna_max_abundance_type"
)

timeseries_rna_max_rel_abund <- make_max_rel_abund_table(
  timeseries_rna_counts,
  value_name = "timeseries_rna_max_rel_abundance",
  type_name = "timeseries_rna_max_abundance_type"
)

timeseries_rna_shared_taxa <- timeseries_rna_active_once |>
  as.data.frame() |>
  select(any_of(colnames(soil_counts_detected))) |>
  colnames()

combined_activity_classes <- bind_rows(
  aquatic_activity_classes,
  shared_activity_classes
) |>
  distinct(OTU, .keep_all = TRUE)

shared_taxa_summary <- tibble(OTU = timeseries_rna_shared_taxa) |>
  left_join(soil_taxa_max_rel_abund, by = "OTU") |>
  left_join(transect_dna_max_rel_abund, by = "OTU") |>
  left_join(transect_rna_max_rel_abund, by = "OTU") |>
  left_join(timeseries_dna_max_rel_abund, by = "OTU") |>
  left_join(timeseries_rna_max_rel_abund, by = "OTU") |>
  left_join(combined_activity_classes, by = "OTU")
```

## Soil versus aquatic transect maximum abundance

```{r plot-soil-vs-transect-max-abundance}
shared_taxa_summary |>
  ggplot(aes(
    x = soil_max_rel_abundance,
    y = transect_dna_max_rel_abundance,
    color = activity_pattern
  )) +
  geom_hline(yintercept = rare_common_threshold, color = "grey") +
  geom_vline(xintercept = rare_common_threshold, color = "grey") +
  geom_point(alpha = 0.25, size = 3) +
  labs(
    x = "Max. Rel. Abundance\nin Soil",
    y = "Max. Rel. Abundance\nin Aquatic Transect"
  ) +
  annotate("text", x = 10^(-1), y = 10^(-1), label = "DNA", size = 6, adj = 1) +
  scale_x_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  scale_y_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  coord_fixed(ratio = 1) +
  scale_color_manual(values = activity_colors, na.value = "grey70") +
  theme(
    legend.position = c(0.98, 0.7),
    legend.justification = c(1, 0),
    legend.box.margin = margin(t = 0, r = 0, b = 0, l = 0),
    legend.text = element_text(face = "bold"),
    legend.background = element_rect(linewidth = 0.3, colour = "black"),
    strip.text = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 14),
    plot.title = element_text(face = "bold", size = 14)
  )

ggsave(filename = "newfigure_dna.pdf", path = paths$figures, width = 6, height = 6, units = "in")

shared_taxa_summary |>
  ggplot(aes(
    x = soil_max_rel_abundance,
    y = transect_rna_max_rel_abundance,
    color = activity_pattern
  )) +
  geom_hline(yintercept = rare_common_threshold, color = "grey") +
  geom_vline(xintercept = rare_common_threshold, color = "grey") +
  geom_point(alpha = 0.25, size = 3) +
  labs(
    x = "Max. Rel. Abundance\nin Soil",
    y = "Max. Rel. Abundance\nin Aquatic Transect"
  ) +
  annotate("text", x = 10^(-1), y = 10^(-1), label = "RNA", size = 6, adj = 1) +
  scale_x_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  scale_y_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  coord_fixed(ratio = 1) +
  scale_color_manual(values = activity_colors, na.value = "grey70") +
  theme(
    legend.position = c(0.98, 0.7),
    legend.justification = c(1, 0),
    legend.box.margin = margin(t = 0, r = 0, b = 0, l = 0),
    legend.text = element_text(face = "bold"),
    legend.background = element_rect(linewidth = 0.3, colour = "black"),
    strip.text = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 14),
    plot.title = element_text(face = "bold", size = 14)
  )

ggsave(filename = "newfigure_rna.pdf", path = paths$figures, width = 6, height = 6, units = "in")
```

## Soil versus aquatic time-series maximum abundance

```{r plot-soil-vs-timeseries-max-abundance}
shared_taxa_summary |>
  ggplot(aes(
    x = soil_max_rel_abundance,
    y = timeseries_dna_max_rel_abundance,
    color = activity_pattern
  )) +
  geom_hline(yintercept = rare_common_threshold, color = "grey") +
  geom_vline(xintercept = rare_common_threshold, color = "grey") +
  geom_point() +
  labs(
    x = "Max. Rel. Abundance\nin Soil",
    y = "Max. Rel. Abundance\nin Aquatic Time Series"
  ) +
  annotate("text", x = 10^(-1), y = 10^(-1), label = "DNA", size = 6, adj = 1) +
  scale_x_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  scale_y_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  coord_fixed(ratio = 1) +
  scale_color_manual(values = activity_colors, na.value = "grey70")

shared_taxa_summary |>
  ggplot(aes(
    x = soil_max_rel_abundance,
    y = timeseries_rna_max_rel_abundance,
    color = activity_pattern
  )) +
  geom_hline(yintercept = rare_common_threshold, color = "grey") +
  geom_vline(xintercept = rare_common_threshold, color = "grey") +
  geom_point() +
  labs(
    x = "Max. Rel. Abundance\nin Soil",
    y = "Max. Rel. Abundance\nin Aquatic Time Series"
  ) +
  annotate("text", x = 10^(-1), y = 10^(-1), label = "RNA", size = 6, adj = 1) +
  scale_x_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  scale_y_log10(limits = c(0.0001, 0.1), labels = scales::label_log()) +
  coord_fixed(ratio = 1) +
  scale_color_manual(values = activity_colors, na.value = "grey70")
```

